\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\title{\textbf{NeuroSAT: Neuro-symbolic Approaches to the SAT Problem}\\[2mm]
\large Reinforcement learning (Alpha(Go)Zero / DQN) with CNNs and Graph Neural
Networks for SAT-solver branching heuristics}
\author{Donato Meoli}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We study two neuro-symbolic approaches that learn branching heuristics for a
Boolean SAT solver: (i) an \emph{Alpha(Go)Zero}-style method that casts SAT
solving as a single-player game and combines Monte Carlo Tree Search (MCTS) with
a convolutional policy/value network~\cite{wang2018}, and (ii)
\emph{Graph-Q-SAT}, a value-based reinforcement-learning method that uses a Graph
Neural Network (GNN) to approximate the $Q$-function over a graph representation
of the formula~\cite{kurin2020}. We additionally introduce \emph{GAT-Q-SAT}, a
variant that injects Graph Attention layers into the Graph-Q-SAT message passing.
The original implementations are several years old and depend on
TensorFlow~1.x, GSL and pinned, hard-to-install libraries. We port both to a
modern, self-contained PyTorch stack, remove the brittle native dependencies, and
\emph{reproduce the original evaluation results exactly}. Our experiments confirm
that the attention-augmented model (GAT-Q-SAT) is strongest on structured
problems (graph colouring), where a latent sub-structure can be inferred, while
on uniform-random 3-SAT near the phase transition the plain Graph-Q-SAT is
preferable.
\end{abstract}

\section{Introduction}
Boolean satisfiability (SAT) is the canonical NP-complete problem and underlies
applications in verification, planning and synthesis. Modern solvers are based on
Conflict-Driven Clause Learning (CDCL) and rely on hand-crafted heuristics ---
most notably VSIDS --- to decide which variable to branch on next. The branching
heuristic is a natural target for machine learning: a good policy can prune the
search tree, especially during the ``warm-up'' phase where counter-based
heuristics make poor decisions.

This project implements and modernises two complementary learning approaches,
developed in the context of the Pattern Recognition and Reinforcement Learning
courses at the University of Pisa:
\begin{itemize}
  \item \textbf{Reinforcement Learning (Alpha(Go)Zero / MCTS)}, following Wang
        and Rompf~\cite{wang2018};
  \item \textbf{Supervised + Reinforcement Learning (GNN + DQN)}, following
        Kurin et al.~\cite{kurin2020}, extended here with a Graph Attention
        Network (GAT) variant~\cite{velickovic2018}.
\end{itemize}

\section{Background}
\paragraph{SAT and CDCL.} A CDCL solver repeatedly (i) \emph{decides} a variable
and assigns it a value, (ii) performs unit propagation, and (iii) on a conflict,
\emph{learns} a clause and backtracks. The branching heuristic chooses the
decision variable; replacing or guiding it with a learned policy is the goal of
both methods studied here.

\paragraph{Reinforcement learning.} We model the problem as a Markov Decision
Process. Graph-Q-SAT uses DQN to approximate the optimal action-value function
$Q^\star(s,a)=\mathbb{E}[r+\gamma\max_{a'}Q^\star(s',a')]$ with a target network
and experience replay. The Alpha(Go)Zero method instead learns a policy prior
$\pi$ and a value $v$, and uses MCTS to obtain improved targets by simulation.

\section{Method 1 --- Alpha(Go)Zero for SAT}
\paragraph{Formulation~\cite{wang2018}.} The CNF formula is represented as a fixed
size sparse matrix of \emph{clauses} $\times$ \emph{variables}, where an entry
encodes the literal polarity ($(1,0)$ for a positive and $(0,1)$ for a negative
occurrence). A simple convolutional network with a policy head $\pi$ (over the
$2\cdot n_{\text{var}}$ actions ``assign variable true/false'') and a value head
$v\in[-1,1]$ scores states. MiniSat is extended into a game environment that
supports MCTS \emph{simulations}: it returns the would-be state for a sequence of
branching decisions without irreversibly changing the real state. Self-play
collects $(\text{state},\ \text{MCTS visit distribution},\ \text{outcome})$
tuples, on which the network is trained with the AlphaZero objective
\begin{equation}
  \mathcal{L} = \underbrace{-\textstyle\sum_a \pi^{\text{MCTS}}_a \log \pi_a}_{\text{cross-entropy}}
              + \underbrace{(z - v)^2}_{\text{value}}
              + c\,\lVert\theta\rVert_2^2 .
\end{equation}
A fixed-size CNN cannot generalise across problem sizes; this limitation is what
motivates the graph-based method below.

\paragraph{Modernisation (this work).} The original code is TensorFlow~1.x and
its C++ environment depends on GSL. We re-implement the three networks
(\texttt{model}, \texttt{model2}, \texttt{model3}) in PyTorch
(\texttt{models\_torch.py}), reproduce the AlphaZero loss and training loop in an
\texttt{AZTrainer} (\texttt{alphazero\_torch.py}), and rewrite the self-play /
supervised-training orchestration (\texttt{main\_torch.py}). Crucially, we
\textbf{remove the GSL dependency}: the Dirichlet exploration noise is
reimplemented with the C++11 \texttt{<random>} facilities (normalised
$\mathrm{Gamma}(\alpha_i,1)$ samples), so the simulation environment builds with
nothing but \texttt{g++} and \texttt{zlib}. The pipeline runs end-to-end on
modern Python/PyTorch and on uniform-random 3-SAT (\texttt{uf20-91}).

\section{Method 2 --- Graph-Q-SAT and GAT-Q-SAT}
\paragraph{Formulation~\cite{kurin2020}.} The state is a bipartite graph with
variable nodes and clause nodes; node features are 2-D one-hot vectors and edge
features encode literal polarity, with an empty global attribute. There are two
actions per unassigned variable; the agent takes the $\arg\max$ of the
$Q$-values over variable nodes. The reward is a constant $-p$ per non-terminal
step and $0$ at termination, encouraging short episodes without reward shaping.
The $Q$-function is an \emph{encode-process-decode} graph network; training uses
DQN. Performance is measured by the Median Relative Iteration Reduction (MRIR):
the ratio of MiniSat iterations to the model's iterations (values $>1$ mean fewer
iterations than MiniSat).

\paragraph{GAT-Q-SAT (this work).} We add an attention pathway to the core graph
block: two \texttt{GATConv} layers (with edge features and configurable heads)
refine the node embeddings before the global update. Attention lets the model
weight neighbouring clauses/variables, which is beneficial precisely when the
instance has an exploitable sub-structure.

\paragraph{Modernisation (this work).} We port Graph-Q-SAT to the latest
PyTorch / PyTorch~Geometric. We drop the hard-to-install
\texttt{torch-scatter}/\texttt{torch-sparse} (replaced by
\texttt{torch\_geometric.utils.scatter}), make the gym dependency
version-robust (the environment is constructed directly, bypassing the
\texttt{gym.make} env-checker, and works on \texttt{gymnasium}), and add a
version-agnostic checkpoint loader that bridges the \texttt{GATConv}
\texttt{lin\_src}/\texttt{lin\_dst}~$\leftrightarrow$~\texttt{lin} naming across
PyTorch~Geometric versions. The stack now uses the latest \texttt{numpy} ($\geq
2$); the MiniSat gym extension is rebuilt against it.

\section{Experiments}
\paragraph{Reproduction (semantic non-regression).} We verified that the
modernised stack does not change the semantics by re-running the evaluation of
the \emph{original trained checkpoints} and comparing against the committed logs.
The numbers match \textbf{exactly}, e.g.\ on the graph-colouring set
\texttt{flat30-60} with a 500-decision cap:
\begin{center}
\begin{tabular}{lccc}
\toprule
Model & MRIR (median) & mean & committed \\
\midrule
Graph-Q-SAT & 1.833 & 2.009 & 1.833 \\
GAT-Q-SAT   & 1.667 & 1.860 & 1.667 \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Graph colouring (MRIR, mean over runs).} GAT-Q-SAT remains strong and
stable across decision caps, including on the larger instances; plain Graph-Q-SAT
degrades on large instances at high caps (dropping below $1$).
\begin{center}
\begin{tabular}{lcccccc}
\toprule
dataset & max10 & max50 & max100 & max300 & max500 & max1000 \\
\midrule
\multicolumn{7}{l}{\emph{GAT-Q-SAT}}\\
flat30-60   & 1.83 & 1.83 & 1.83 & 1.83 & 1.83 & 1.83 \\
flat100-239 & 1.99 & 1.99 & 1.99 & 1.98 & 1.98 & 1.98 \\
flat150-360 & 2.24 & 2.32 & 2.28 & 2.17 & 2.14 & 2.12 \\
flat200-479 & 2.12 & 2.11 & 2.15 & 1.99 & 1.90 & 1.79 \\
\midrule
\multicolumn{7}{l}{\emph{Graph-Q-SAT}}\\
flat30-60   & 1.32 & 1.13 & 1.13 & 1.13 & 1.13 & 1.13 \\
flat100-239 & 1.87 & 1.79 & 1.69 & 1.61 & 1.60 & 1.60 \\
flat150-360 & 2.06 & 1.75 & 1.61 & 1.28 & 1.20 & 1.18 \\
flat200-479 & 1.90 & 1.95 & 1.68 & 1.37 & 1.10 & 0.83 \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Uniform-random 3-SAT.} The picture reverses: on random instances near
the phase transition (clause/variable ratio $\approx 4.3$) the attention model is
only mediocre ($\sim 1.2$--$1.8$ MRIR), whereas plain Graph-Q-SAT is markedly
stronger (e.g.\ up to $\sim 5\times$ on \texttt{uf250-1065}). This supports the
interpretation that attention helps when there is a structure to infer (graph
colouring) and not on near-uniform random formulas.

The tables above are generated directly from the evaluation logs
(\texttt{GQSAT/runs/*/*.tsv}) by \texttt{aggregate\_results.py}, replacing the
previous manual spreadsheet workflow.

\section{Related Work}
The field has converged on graph- and attention-based models for SAT. NeuroSAT
learns satisfiability from single-bit supervision~\cite{neurosat}; NeuroCore
predicts unsat-core variables to refine branching~\cite{neurocore}; NeuroComb and
\textbf{NeuroBack}~\cite{neuroback} move inference \emph{offline} (a single query
before solving) and, using a Graph Transformer with self-attention, achieve the
first wall-clock speed-ups on industrial instances (Kissat, SAT Competition).
This trajectory is notable for two reasons relevant to this project: (i) the move
to \emph{attention} on SAT graphs corroborates the GAT-Q-SAT direction explored
here; and (ii) the known limitation of Graph-Q-SAT --- online inference is too
slow for wall-clock gains --- is exactly what the offline, one-shot inference of
NeuroBack overcomes. More recent work continues along the RL-from-algorithm-feedback
and restricted-heuristics lines~\cite{mlforsat2023}, and surveys catalogue the
design space~\cite{survey2025}.

\section{Conclusions}
We modernised two neuro-symbolic SAT methods to a single, self-contained,
latest-version PyTorch stack, removed their brittle native dependencies
(TensorFlow~1.x, GSL, \texttt{torch-scatter}/\texttt{torch-sparse}, pinned gym),
and reproduced the original results exactly. The experiments confirm the central
qualitative finding: graph attention is advantageous on structured problems
(graph colouring) and not on uniform-random SAT. The Alpha(Go)Zero CNN method,
while elegant, is fundamentally limited by its fixed-size input and does not
generalise across sizes --- the gap that the graph-based methods, and the
subsequent literature, were designed to close.

\begin{thebibliography}{9}
\bibitem{wang2018} F. Wang and T. Rompf.
  \emph{From Gameplay to Symbolic Reasoning: Learning SAT Solver Heuristics in
  the Style of Alpha(Go) Zero}. arXiv:1802.05340, 2018.
\bibitem{kurin2020} V. Kurin, S. Godil, S. Whiteson, B. Catanzaro.
  \emph{Can Q-Learning with Graph Networks Learn a Generalizable Branching
  Heuristic for a SAT Solver?} NeurIPS 2020. arXiv:1909.11830.
\bibitem{velickovic2018} P. Veli\v{c}kovi\'c et al.
  \emph{Graph Attention Networks}. ICLR 2018. arXiv:1710.10903.
\bibitem{neurosat} D. Selsam et al.
  \emph{Learning a SAT Solver from Single-Bit Supervision}. arXiv:1802.03685.
\bibitem{neurocore} D. Selsam and N. Bj{\o}rner.
  \emph{Guiding High-Performance SAT Solvers with Unsat-Core Predictions}.
  arXiv:1903.04671.
\bibitem{neuroback} W. Wang et al.
  \emph{NeuroBack: Improving CDCL SAT Solving using Graph Neural Networks}.
  ICLR 2024. arXiv:2110.14053.
\bibitem{mlforsat2023} \emph{Machine Learning for SAT: Restricted Heuristics and
  New Graph Representations}. arXiv:2307.09141, 2023.
\bibitem{survey2025} \emph{Neural Approaches to SAT Solving: Design Choices and
  Interpretability}. arXiv:2504.01173, 2025.
\end{thebibliography}

\end{document}
